Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-03-16 09.0_manuscript.rmd
Updated mortality results

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and , respectively; Fig. S11). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017]. Species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) non-linearly declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and experienced greater hydraulic-risk, which increased over 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time)). For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig. \@ref(fig:erd-time)), but during the dry season (Fig. \@ref(fig:swp-droughts)), and without elevated mortality rates (Fig. \@ref(fig:erd-mort-census-evg)). Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).

>>>>

Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with an El Niño event identiﬁed based on Oceanic Niño Index values > 0.5 °C for at least five consecutive months [@Detto.2018; @Condit.2017]. Out of those six intervals, among evergreen species, ERD explained 30-40% variation in mortality rates during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values significant at alpha = 0.05 for four intervals and at alpha = 0.1 for one interval; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained 10-16% variation in mortality in four intervals with El-Nino events (p-values marginally significant, at alpha = 0.1; Fig. S11).

Our analysis of hydrological droughts ranged from 1990 to 2015 and revealed distinctive extreme hydrological droughts for the census-intervals 1995-2000, 2000-2005 and 2005-2010 for which ERD explained significant mortality (Fig. \@ref(fig:swp-droughts)). Census interval 1995-2000 was marked with a prolonged and extreme hydrological drought during the 1997-1998 El-Nino event. During 2000-2005 and 2005-2010 extreme hydrological drought occurred in the wet season, effectively extending the dry season into the wet season (Fig. 3). While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during census-intervals 1982-1985 and 1985-1990, for which ERD explained significant mortality, BCI experienced one and three extreme deficit water years, respectively; and with the El-Niño of 1983 marked by the harshest dry season since 1931 [@Condit.2017].

On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which steadily increased over 1995-2000, 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time)). For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig. \@ref(fig:erd-time)), but without elevated mortality rates (Fig. \@ref(fig:erd-mort-census-evg)). Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
